---
title: "Recovering a spherical Gaussian field with the MRTS basis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recovering a spherical Gaussian field with the MRTS basis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  dpi = 90
)
set.seed(1)
```

## What this vignette shows

Given noisy point observations on the sphere, we use the multi-resolution
thin-plate spline (MRTS) basis from `mrts_sphere()` to reconstruct the
underlying field. The simulated field is a spherical Gaussian process
with exponential covariance in great-circle distance — built using
`fields::rdist.earth()` rather than any internal helper, so the example
is fully transparent.

```{r}
library(mrtsSphere)
```

## A global lat/lon grid

```{r}
n_lon <- 24
n_lat <- 12
lon_seq <- seq(-180, 165, length.out = n_lon)
lat_seq <- seq( -85,  85, length.out = n_lat)
grid <- as.matrix(expand.grid(lat = lat_seq, lon = lon_seq))
nrow(grid)
```

## Simulating a spherical Gaussian field

We compute a great-circle distance matrix and use it to build an
exponential covariance matrix. Then a single draw from `N(0, Sigma)`
gives the "true" field on the grid.

```{r}
# fields::rdist.earth expects (lon, lat); our grid is (lat, lon).
d_grid <- fields::rdist.earth(grid[, 2:1], miles = FALSE)
range_km   <- 2000
sigma2     <- 1
cov_field  <- sigma2 * exp(-d_grid / range_km)

# Draw via Cholesky (stable jitter for numerical PSD).
L      <- t(chol(cov_field + diag(1e-8, nrow(grid))))
y_true <- as.numeric(L %*% rnorm(nrow(grid)))
```

## Sampling noisy observations

```{r}
n_obs    <- 80
obs_idx  <- sample(nrow(grid), n_obs)
obs_loc  <- grid[obs_idx, ]
y_obs    <- y_true[obs_idx] + rnorm(n_obs, sd = 0.3)
```

## Building the MRTS basis

`mrts_sphere(knot, k, X)` evaluates `k` MRTS basis functions, defined
on the sphere from the knot set, at the rows of `X`. We use the
observation locations as knots and evaluate everywhere on the prediction
grid.

```{r}
k <- 12
basis <- mrts_sphere(knot = obs_loc, k = k, X = grid)$mrts
dim(basis)
```

## Fitting and predicting

Project the noisy observations onto the basis (least squares), then
predict on the full grid.

```{r}
B_obs    <- basis[obs_idx, , drop = FALSE]
beta_hat <- qr.solve(B_obs, y_obs)
y_hat    <- as.numeric(basis %*% beta_hat)

rmse <- sqrt(mean((y_hat - y_true)^2))
round(rmse, 3)
```

## Visualising

```{r, fig.alt = "Three lat/lon maps showing the truth, observations, and MRTS prediction."}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))

zlim <- range(y_true, y_hat)

image(matrix(y_true, n_lat, n_lon),
      x = lat_seq, y = lon_seq,
      xlab = "lat", ylab = "lon",
      main = "truth", zlim = zlim)

plot(obs_loc[, "lon"], obs_loc[, "lat"],
     xlim = range(lon_seq), ylim = range(lat_seq),
     xlab = "lon", ylab = "lat",
     main = sprintf("observations (n = %d)", n_obs),
     pch = 16, cex = 0.8)

image(matrix(y_hat, n_lat, n_lon),
      x = lat_seq, y = lon_seq,
      xlab = "lat", ylab = "lon",
      main = sprintf("MRTS prediction (k = %d)", k), zlim = zlim)

par(op)
```

## What just happened

The first MRTS basis function is the constant `sqrt(1/n_knot)` and
captures the field's mean. Subsequent basis functions come from the
eigen-decomposition of the centered spherical kernel evaluated at the
knot set, ordered by descending eigenvalue, so they capture progressively
finer-scale spherical structure. With only `r 12` basis functions and
`r 80` noisy observations we recover most of the smooth field; pushing
`k` higher would let the basis express finer structure at the cost of
more variance.

## A more principled fit: MRMM

The OLS fit above ignores residual spatial correlation. The
**multi-resolution mixed model** ([`mrmm_fit()`]) adds a tapered Matérn
residual covariance and fits the basis coefficients by GLS, then
delivers kriging-style predictions with a closed-form prediction
standard error.

You supply five covariance parameters:

- `taper_range` — Wendland compact-support range (radians on the unit
  sphere)
- `matern_range` — Matérn range
- `sill`, `nugget` — variance components
- `smoothness` — Matérn smoothness \(\nu\) (default `0.5`, exponential)

```{r mrmm-fit}
fit <- mrmm_fit(
    z            = y_obs,
    loc          = obs_loc,
    knots        = obs_loc,
    k            = 12,
    taper_range  = 1.0,
    matern_range = 0.3,
    sill         = 0.6,
    nugget       = 0.1,
    smoothness   = 0.5
)
fit
```

Predict on the full grid with mean and prediction SE:

```{r mrmm-predict}
pred <- predict(fit, newdata = grid, se.fit = TRUE)
str(pred)
sqrt(mean((pred$fit - y_true)^2))   # RMSE — typically lower than the OLS fit
```

```{r mrmm-plot, fig.alt = "Three lat/lon maps showing truth, MRMM mean prediction, and MRMM SE."}
op <- par(mfrow = c(1, 3), mar = c(2, 2, 2, 1))

zlim <- range(y_true, pred$fit)

image(matrix(y_true,    n_lat, n_lon), x = lat_seq, y = lon_seq,
      xlab = "lat", ylab = "lon", main = "truth", zlim = zlim)

image(matrix(pred$fit,  n_lat, n_lon), x = lat_seq, y = lon_seq,
      xlab = "lat", ylab = "lon", main = "MRMM mean", zlim = zlim)

image(matrix(pred$se.fit, n_lat, n_lon), x = lat_seq, y = lon_seq,
      xlab = "lat", ylab = "lon", main = "MRMM SE")

par(op)
```

In v0.2.0 the covariance parameters must be supplied. A
parameter-estimation routine (`mrmm_estimate_params()`) is planned for a
future release; in the meantime users can fit the parameters separately
with `geoR`, `gstat`, or the variogram-based code in
`inst/paper/fn_0610.R`, then plug them into `mrmm_fit()`.

## Reference

Huang, H.-Y., Huang, H.-C., and Ing, C.-K. (2025). Multi-Resolution
Spatial Methods on the Sphere: Efficient Prediction for Global Data.
*Environmetrics*. <https://doi.org/10.1002/env.70092>
